---
title: "Data Collection and Missing Data"
format: 
  revealjs:
    theme: default
    chalkboard: true
    slide-number: true
---

```{r, echo=FALSE, warning=FALSE}
library(knitr)
library(ggplot2)
library(kableExtra)
library(patchwork)
library(tidybayes)
library(cmdstanr)
check_cmdstan_toolchain(fix = TRUE, quiet = TRUE)
register_knitr_engine(override = FALSE)
options(mc.cores=4)
suppressPackageStartupMessages(library(tidyverse))
opts_chunk$set(echo=FALSE, fig.align='center', out.width="60%") 
knitr::opts_chunk$set(dpi=300, fig.width=7)
options("kableExtra.html.bsTable" = T)
ig <- function(file) {knitr::include_graphics(file)}
source("../scripts/make_figs.R")
source("../scripts/color_defs.R")
```

### Announcements

- Chapter 8 and 18

## Notation

- $y = (y_1, . . . , y_N )$, y is a matrix ($y_i$ can be a $p$-vector.

- $I = (I_1, . . . , I_N)$ be a matrix of the same dimensions as y 

-  ‘obs’= $\{(i,j): I_{ij} = 1\}$ and ‘mis’= $\{(i, j) : I_ij = 0\}$

## Ignorability



## Ignorability

- Two conditions imply ignorability: 
  + Missing at random
  + Distinct parameters
  
## Censoring and Truncation

```{r}
library(cmdstanr)
register_knitr_engine(override = TRUE)
```


```{r}

truncation_point <- 1/2 ## Phi
ntotal <- 1000
true_mean <- 1
y <- rnorm(ntotal, mean=true_mean, sd=1)
nobs <- sum(y < truncation_point)
yobs <- y[y < truncation_point]
ybar_obs <- mean(yobs)
```


```{r}
ggplot(tibble(y=y, obs=y < truncation_point)) + geom_histogram(aes(x=y, alpha=obs)) +
  scale_alpha_discrete(range = c(0.25, 1)) +
  theme_bw() + 
  geom_vline(xintercept=ybar_obs, col="blue") + 
  geom_vline(xintercept=true_mean, col="red") + 
  ggtitle('N=50, theta=0, phi=0.5')
```

## Censoring (known point)

## Censoring (known point)
```{stan, output.var="cens_known", engine=cmdstan, echo=TRUE}
data {

  real<lower=0> truncation_point;
  int<lower=0> ntotal;
  int<lower=0> nobs;
  real ybar_obs;
} parameters {
  real theta;
}
model {
  ybar_obs ~ normal(theta, 1.0/sqrt(nobs));
  target += (ntotal-nobs)*normal_lccdf(truncation_point | theta, 1.0);

}
```

## {.smaller}
```{r}
fit <- cens_known$sample(data=list(ybar_obs=ybar_obs, ntotal = ntotal, nobs=nobs, truncation_point=0.5), refresh=100, show_messages=FALSE, show_exceptions=FALSE)
fit$summary()
```

## Censoring (unknown point)

## Censoring (unknown point)
```{stan, output.var="cens_unknown",  engine=cmdstan, echo=TRUE}
data {
  real ybar;
  real<lower=ybar> ymax;
  real<lower=0> ntotal;
  real<lower=0> nobs;
} parameters {
  real theta;
  real<lower=ymax> truncation_point;
}
model {
  target += normal_lpdf(ybar | theta, 1.0/sqrt(nobs)) +
  (ntotal-nobs)*normal_lccdf(truncation_point | theta, 1.0);
}
```


```{r, message=FALSE}
fit_unknown <- cens_unknown$sample(data=list(ybar=ybar_obs, ymax = max(yobs), ntotal=ntotal, nobs=nobs), refresh=10, show_messages=FALSE, show_exceptions=FALSE)
```

## Unknown censoring point{.smaller}

```{r}
fit_unknown$summary()
```

## Truncation (known point)

```{stan, output.var="trunc_known", engine=cmdstan, echo=TRUE}
data {
  real ybar;
  real<lower=ybar> ymax;
  real<lower=0> nobs;
  real<lower=ymax> truncation_point;
} parameters {
  real theta;
}
model {
  target += normal_lpdf(ybar | theta, 1.0/sqrt(nobs)) -
  nobs*log(1.0 - Phi(theta - truncation_point));
}
```

## Truncation (known point) {.smaller}
```{r, message=FALSE}
fit_unknown <- trunc_known$sample(data=list(ybar=ybar_obs, ymax = max(yobs), ntotal=ntotal, nobs=nobs, truncation_point=truncation_point), refresh=0, show_messages=FALSE, show_exceptions=FALSE)
```

```{r}
fit_unknown$summary()
```

## Truncation (unknown point)

## Truncation (unknown point)
```{stan, output.var="trunc_unknown",  engine=cmdstan, echo=TRUE}
data {
  real ybar;
  real<lower=ybar> ymax;
  real<lower=0> ntotal;
  real<lower=0> nobs;
} parameters {
  real theta;
  real<lower=ymax> truncation_point;
}
model {
  truncation_point ~ normal(0, 10);
  target += normal_lpdf(ybar | theta, 1.0/sqrt(nobs)) -
  nobs*log(1.0 - Phi(theta - truncation_point));
}
```

## Truncation (unknown point) {.smaller}
```{r, message=FALSE}
fit_unknown <- trunc_unknown$sample(data=list(ybar=ybar_obs, ymax = max(yobs), ntotal=ntotal, nobs=nobs), refresh=0, show_messages=FALSE, show_exceptions=FALSE)
```

```{r}
fit_unknown$summary()
```

## Multivariate Normal


```{r}
library(MASS)

# Simulate some data
mu <- c(10, 5, -5)
Sig <- matrix(c(1, 0.7, 0.8,
                0.7, 2, 0.2,
                0.8, 0.2, 1.5), 
              3, 3)
N <- 1000
y <- mvrnorm(N, mu, Sig)

# Randomly remove some data (MAR)
nmiss <- 315
miss <- sample.int(prod(dim(y)), size = nmiss)
y[miss] <- NA
```


```{r}
# Extract the missing values into a VECTOR
yobs <- y[!is.na(y)]

# Extract the missing and present values as MATRICES
ind_obs <- which(!is.na(y), arr.ind = TRUE)
ind_miss <- which(is.na(y), arr.ind = TRUE)
```

## MVN with missingness in STAN

```{stan, output.var="mvn_miss", engine = "cmdstan"}
data {
    int<lower=0> Nrow;
    int<lower=0> Ncol;

    int<lower=0> Nobs; // Number of non-missing values
    int<lower=0> Nmiss; // Number of missing values

    array[Nobs] real yobs;   // Vector of non-missing values
    array[Nobs, 2] int ind_obs;     // Matrix (row, col) of non-missing value indices
    array[Nmiss, 2] int ind_miss;     // Matrix (row, col) of missing value indices
}
parameters {
    // Multivariate normal distribution parameters
    vector[Ncol] mu;
    cov_matrix[Ncol] Sigma;

    // Vector containing "stochastic" nodes (for filling missing values
    array[Nmiss] real ymiss;      
}
transformed parameters {
    array[Nrow] vector[Ncol] y;   // The "data" with interpolated missing values
    // Fill y with non-missing values 
    for(n in 1:Nobs) {
        y[ind_obs[n,1]][ind_obs[n,2]] = yobs[n];
    }
    // Fill the rest of y with missing value "parameters"
    for(n in 1:Nmiss){
        y[ind_miss[n,1]][ind_miss[n,2]] = ymiss[n];
    }
}
model {
    for(i in 1:Nrow){
        target += multi_normal_lpdf(y[i] | mu, Sigma);
    }
}
```

```{r}
mvn_data <- list(Nrow = nrow(y),
                 Ncol = ncol(y),
                 Nobs = length(yobs),
                 Nmiss = sum(is.na(y)),
                 yobs = yobs,
                 ind_obs = ind_obs,
                 ind_miss = ind_miss)

fit <- mvn_miss$sample(data = mvn_data, refresh=0)

fit %>% spread_draws(Sigma[i, j]) %>% 
  left_join(expand_grid(i = 1:3, j=1:3) %>% mutate(true = as.numeric(Sig))) %>% 
  ggplot() + geom_histogram(aes(x=Sigma)) + 
  geom_vline(aes(xintercept=true), col="red") + 
  facet_wrap(~i + j) + theme_bw() + ggtitle("Sigma")
```



            